Fasano--Franceschini Test: an Implementation of a 2-Dimensional Kolmogorov--Smirnov in \proglang{R}

### 2021-06-08 journalMS/article.tex
RB edits finally

%% - \Abstract{} almost as usual
\Abstract{
\RBnote{Are citations allowed in the abstract (per journal style)? - Yes, citation are found in the abstract both in the style guide template and also from checking published articles }
The versatility of the univariate Kolmogorov-Smirnov (KS) test has made it a cornerstone of statistical analysis and is commonly used across the scientific disciplines. Nonetheless, a generalization of the KS test as proposed by Kolmogorov and Smirnov does not naturally extend when working with data in more than one dimension. Here, we present the \pkg{fasano.franceschini.test} package, an  \proglang{R} implementation of the 2-D KS two--sample test as defined by Fasano and Franceschini~\citep{Fasano1987}. The \pkg{fasano.franceschini.test} package provides three improvements over the current 2-D KS test on the Comprehensive \proglang{R} Archive Network (CRAN): (i) the Fasano and Franceschini test has been shown to run in $O(n^2)$ versus the Peacock implementation which runs in $O(n^3)$; (ii) the package implements a procedure for handling ties in the data; and (iii) the package implements a parallelized bootstrapping procedure for improved significance testing. Ultimately, the \pkg{fasano.franceschini.test} package presents a robust statistical test for analysing random samples defined in 2-dimensions.
\RBnote{Summarize what the paper will cover. - ENC: Rewrote abstract}
}

@@

% R -> Peacock test (computationally inificent)
% Python <- Implementation
% C <- Implementation

\RBnote{Please use nonbreaking spaces (tilde rather than space) before all cite commands - ENC: tilde has be added before each cite, removing all spaces}

The Kolmogorov--Smirnov (KS) is a non--parametric, univariate statistical test designed to assess whether a set of data is consistent with a given probability distribution. First derived by Kolmogorov and Smirnov in a series of papers~\citep{Kolmogorov1933,Kolmogorov1933a,Smirnov1936,Smirnov1937,Smirnov1939,Smirnov1944,Smirnov1948}, the one-sample KS test defines the distribution of the quantity $D_{KS}$, the maximal absolute difference between the empirical cumulative distribution function (CDF) of a set of values and a reference probability distribution. Kolmogorov and Smirnov's key insight was proving the distribution of $D_{KS}$ was independent of the CDFs being tested. Thus, the test can effectively be used to compare any univariate empirical data distribution to any continuous univariate reference distribution. A two-sample KS test could further be used to compare any two univariate empirical data distributions against each other to determine if they are drawn from the same underlying univariate distribution.

The versatility of the univariate KS test has made it a cornerstone of statistical analysis and is commonly used across the scientific disciplines~\citep{Atasoy2017,Chiang2018,Hahne2018,Hargreaves2020,Wong2020,Kaczanowska2021} \RBnote{ENC: Cited a broad range of fields (neuropharmacology, climate change, covid-19, sustainable nutrition, robotic prostetics, cancer therapeutics)}. Nonetheless, a generalization of the KS test as proposed by Kolmogorov and Smirnov does not naturally extend when working with data in more than one dimension. Fortunately, a solution to the dimensionality issue was articulated by Peacock~\citep{Peacock1983} and later extended by Fasano and Franceschini~\citep{Fasano1987}.

Currently, only the Peacock implementation of the 2-D two-sample KS test is available in \proglang{R}~\citep{R} with the \pkg{Peacock.test} package via the \fct{peacock2} function, but this has been shown to be markedly slower than the Fasano and Franceschini algorithm~\citep{Lopes2007}. A \proglang{C} implementation of the Fasano--Franceschini test is available in~\citep{numericalRecipes}; however, arguments have been made to the validity of the implementation of the test not being distribution-free~\citep{Babu2006}. Furthermore, the \proglang{C} implementation statistical testing is based on a fit to Monte Carlo Simulation and is only valid for significance levels $\alpha \lessapprox 0.20$, which may present potential issues in some analysis.

Here we present the \pkg{fasano.franceschini.test} package as an \proglang{R} implementation of the 2-D two-sample KS test as defined by Fasano and Franceschini~\citep{Fasano1987} The \pkg{fasano.franceschini.test} package provides two improvements over the current 2-D KS test on the Comprehensive \proglang{R} Archive Network (CRAN): (i) the Fasano and Franceschini test has been shown to run in $O(n^2)$ versus the Peacock implementation which runs in $O(n^3)$; and (ii) the package implements a bootstrapping procedure for improved significance testing and mitigates the concerns over the distribution-free nature of the test brought up by~\citep{Babu2006}.

% \begin{leftbar}
% The introduction is in principle ``as usual''. However, it should usually embed
% both the implemented \emph{methods} and the \emph{software} into the respective
% relevant literature. For the latter both competing and complementary software
% should be discussed (within the same software environment and beyond), bringing
% out relative (dis)advantages. All software mentioned should be properly
% \verb|\cite{}|d. (See also Appendix~\ref{app:bibtex} for more details on
% \textsc{Bib}{\TeX}.)
%
% For writing about software JSS requires authors to use the markup
% \verb|\proglang{}| (programming languages and large programmable systems),
% \verb|\pkg{}| (software packages), \verb|\code{}| (functions, commands,
% arguments, etc.). If there is such markup in (sub)section titles (as above), a
% plain text version has to be provided in the {\LaTeX} command as well. Below we
% also illustrate how abbrevations should be introduced and citation commands can
% be employed. See the {\LaTeX} code for more details.
% \end{leftbar}
%
% Modeling count variables is a common task in economics and the social sciences.
% The classical Poisson regression model for count data is often of limited use in
% these disciplines because empirical count data sets typically exhibit
% overdispersion and/or an excess number of zeros. The former issue can be
% addressed by extending  the plain Poisson regression model in various
% directions: e.g., using sandwich covariances or estimating an additional
% dispersion parameter (in a so-called quasi-Poisson model). Another more formal
% way is to use a negative binomial (NB) regression. All of these models belong to
% the family of generalized linear models (GLMs). However, although these models
% typically can capture overdispersion rather well, they are in many applications
% not sufficient for  modeling excess zeros. Since \cite{Mullahy:1986} there is
% increased interest in zero-augmented models that address this issue by a second
% model component capturing zero counts. An overview of count data models in
% econometrics, including  hurdle and zero-inflated models, is provided in
% \cite{Cameron+Trivedi:2013}.
%
% In \proglang{R} \citep{R}, GLMs are provided by the model fitting functions
% \fct{glm} in the \pkg{stats} package and \fct{glm.nb} in the \pkg{MASS} package
% \citep[][Chapter~7.4]{Venables+Ripley:2002} along with associated methods for
% diagnostics and inference. The manuscript that this document is based on
% \citep{Zeileis+Kleiber+Jackman:2008} then introduced hurdle and zero-inflated
% count models in the functions \fct{hurdle} and \fct{zeroinfl} in the \pkg{pscl}
% package \citep{Jackman:2015}. Of course, much more software could be discussed
% here, including (but not limited to) generalized additive models for count data
% as available in the \proglang{R} packages \pkg{mgcv} \cite{Wood:2006},
% \pkg{gamlss} \citep{Stasinopoulos+Rigby:2007}, or \pkg{VGAM} \citep{Yee:2009}.

%% -- Manuscript ---------------------------------------------------------------

%% - In principle "as usual" again.
%% - When using equations (e.g., {equation}, {eqnarray}, {align}, etc.
%%   avoid empty lines before and after the equation (which would signal a new
%%   paragraph.
%% - When describing longer chunks of code that are _not_ meant for execution
%%   (e.g., a function synopsis or list of arguments), the environment {Code}
%%   is recommended. Alternatively, a plain {verbatim} can also be used.
%%   (For executed code see the next section.)

@@

\RBnote{This can be shortened substantially.  I have done so, but please look closely.  In general, avoid writing textbook material in papers.}

The Kolmogorov--Smirnov (KS) test is a non--parametric method for determining whether a sample is consistent with a given probability distribution. %While a comprehensive discussion pertaining the intricacies of the 1-D KS test can be found elsewhere~\citep{Stephens1992a}, we will present a brief overview as motivation for the basis and intuition of the 2-D KS test.
%
%As described previously, the 1-D KS test can be divided into two variants of the same statistical procedure:
%
%\begin{leftbar}
%\textbf{The One-Sample Test}
%\begin{itemize}
%\item Compares the data to analytical distribution.
%\end{itemize}
%\textbf{The Two-Sample Test}
%\begin{itemize}
%\item Compares two data samples to determine if they are drawn from the same underlying probability distribution.
%\end{itemize}
%\end{leftbar}
%
%In both cases, the KS test works as follows:
%
In one dimension,
the Kolmogorov-Smirnov statistic ($D_{KS}$) is the defined by the maximum absolute difference between the cumulative density functions of the data and model (one--sample), or between the two data sets (two--sample),
as illustrated in (\textbf{Figure~\ref{fig:kstest1D}}).
%. In the following example, the orange and blue curve in the left panel represents the underlying distribution (pdf) for sample1 and sample2, respectively (\textbf{Figure \ref{fig:kstest1D}}). The eCDF is then defined by data drawn from these underlying distribution, and the D--stat is the computed as the largest between difference between the eCDF (black dotted line) in the right panel.

\begin{figure}[t!]
\centering
\includegraphics{pdfvsCDF}
\caption{\label{fig:kstest1D} \textbf{LEFT:} Probability density function (PDF) of two normal distribution: orange sample1 $\mathcal{N}(\mu = 0,\,\sigma^{2} = 1)$; blue sample2 $\mathcal{N}(\mu = 5,\,\sigma^{2} = 1)$. \textbf{RIGHT:} Empirical cumulative density functions (eCDF) of the two PDFs in the left panel. Black dotted line represents the maximal absolute difference between the eCDFs ($D_{KS}$).
}
\end{figure}

@@

In the large--sample limit ($n \geq 80$), it can be shown~\citep{Kendall1946} that $D_{KS}$ converges in distribution to
\begin{equation} \label{eq:1}
\Phi(\lambda) = 2 \sum_{k=1}^{\infty} -1^{k-1}e^{-2k^2\lambda^2} \,.
\end{equation}
%Given the maximum absolute difference ($D_{KS}$) and \textbf{Equation~\ref{eq:1}}, one can  compute a $p$-value for  hypothesis testing to determine if there is a significance difference between the cumulative density functions of the data and model (one-sample), or between the two data sets (two-sample).
%
%\textbf{For the 1-Sample case}, the $p$-value probability of a given D is defined by
In the one-sample case, the $p$~value is given by
\begin{equation} \label{eq:2}
\mathbb{P}(D > observed) = \Phi ( D\sqrt{n})\,;
\end{equation}
in the two-sample case, the $p$~value is given by
%where $N$ is the number of data points sampled. A $p$-value $< 0.05$, indicates a significant difference between the samples.
%
%\textbf{In the 2-Sample case}, the $p$-value probability of a given D is defined by
\begin{equation} \label{eq:3}
\mathbb{P}(D > observed) = \Phi \left( D\sqrt{\frac{n_1n_2}{n_1+n_2}} \right)\,.
\end{equation}
%
%where $n_1$ and $n_2$ is the number of data points in the first and second samples respectively. A $p$-value $< 0.05$, indicates a significant difference between the samples.
\RBnote{Note usage of left( and right) [alternatively, or amsmath big( ] to make bigger parens}

\subsection{Higher Dimensional KS Test}

%Similar to the 1-D KS test, the obvious solution to extend the theory into higher dimensional would be to define the maximal cumulative difference between the two 2-D distributions. This however, presents an issue as the cumulative distribution function is not well defined in more than one dimension. In 2-D, there are 4 ways (3 independent ways) of defining the cumulative distribution, since the direction in which we define choose to order the points in the x and y are arbitrary (\textbf{Figure \ref{fig:kstest2Dissue}}). In N dimensional space there are $2^{N}-1$ independent ways of defining the cumulative distribution function~\citep{Peacock1983}.
Extending the above to two or higher dimension is complicated by the fact that CDFs are not well-defined in more than one dimension.  In 2-D, there are 4 ways (3 independent ways) of defining the cumulative distribution, since the direction in which we order the $x$ and $y$ points are arbitrary (\textbf{Figure \ref{fig:kstest2Dissue}}); more generally, in $P$-dimensional space there are $2^{P}-1$ independent ways of defining the cumulative distribution function~\citep{Peacock1983}.

\begin{figure}[t!]
\centering
\includegraphics{CDF2Dissue}
\caption{\label{fig:kstest2Dissue} Four ways (3 independent) of defining the cumulative distribution for a given point in 2-D.
%$P(x < X, y < Y) + P(x < X, y > Y) + P(x > X, y < Y) = 1 - P(x > X, y > Y)$
}
\end{figure}

\subsection{Peacock Test (1983) and Fasano--Franceschini Test (1987)}
Peacock solved the higher dimensionality issue by defining the 2-D test statistic as the largest difference between the empirical and theoretical cumulative distributions, after taking all possible ordering combinations into account. Peacock's test thus computes the total integrated probability -- i.e.\,fraction of data -- in each of the four quadrants around all possible points in the data. For example, for a given $n$ points in a two-dimensional space, the cumulative distribution functions is calculated in the $4n^2$ quadrants of the plane defined by all pairs ($X_i$ ,$Y_j$), $X_i$ and $Y_j$ being coordinates of any pairs of points in the given sample. By ranging over all possible pairs of data points and quadrants, the 2-dimensional $D$ statistic is defined by the maximal difference of the integrated probabilities between samples.

The variation defined by Fasano and Franceschini was to only consider quadrants centered in each point of the given samples to compute the cumulative distribution functions.  Rather than looking over all $X_i, Y_j: i,j={1,n}$, Fasano and Franceschini only used the detected points $X_i, Y_i : i={1,n}$. Thus for any given $n$ points in a two-dimensional space, those $n$ points define $4n$ -- rather than $4n^2$ -- quadrants.

The 2-D two sample Fasano and Franceschini test is illustrated in the following example to compare the 2-D distributions of orange versus blue points (\textbf{Figure \ref{fig:kstest2D}}).

@@

The algorithm loops through each point in the orange sample in turn to define the origin of 4 quadrants (grey dotted line). The fraction of orange and blue points in each quadrant is computed (corners of plot) and the quadrant with the maximal difference is designated with the current maximum for the specified origin. By iterating over all pairs of data points and quadrants, the $D_{FF,1}$ is defined by the maximal difference of the integrated probabilities between samples. In this case using the orange point as the origin, the maximal difference is $D_{FF,1} = 0.52$.
\RBnote{How are tied values counted? Describe that here.  Also discuss the treatment of the origin. - ENC: Discussion Added below}

\RBnote{Since $D_{FF}$ from one sample isn't the final $D_{FF}$, you should note it appropriately (eg, $D_{FF,1}$ and $D_{FF,2}$). - ENC: Text has been updated accordingly}
This process is repeated using the blue points as the origins to compute the maximal $D_{FF,2}$. Finally, both maximal $D_{FF,1}$ and $D_{FF,2}$ statistics using the orange and blue points as origins are averaged to compute the overall $D_{FF}$ for hypothesis testing.

Note that in real data, it is possible for some points to be tied with the X and/or Y coordinates of the origin -- creating an ambuguity when computing the fraction of points in each quadrant. Since the Fransono Francheschini test attempts to define the maximal difference of the integrated probabilities, a natural solution would be to treat a point that is tied with the current X and/or Y coordinates of the origin as equally likely to have been drawn from any of the tied quadrants. In doing so, the integrated probability of the given point is exactly evenly distributed across the tied quadrants. For example, any data point sharing the same X or Y coordianted as the current origin is evenly distributed across the tied quadrants, with each of the two quadrants recieveing half a count. Any data point sharing the same X and Y coordiantes as the current origin  -- the origin included -- is evenly distributed across all quadrants, with all four quadrants recieveing a quarter count.

@@

Using Monte Carlo simulation, Fasano and Franceschini created a look-up table of critical values of $D_{FF}$ as a function of $D_{FF}$, the sample size $N$, and the coefficient of correlation $r$. Teukolsky and Press later defined an approximate fit to this data in equation \ref{eq:4} for quick computation~\cite{numericalRecipes}, as follows.

For the 1-Sample case,
\begin{equation} \label{eq:4}
\mathbb{P}(D > observed) = \Phi \left( \frac{D\sqrt{N}}{1+\sqrt{1-r^2}(0.25-0.75/\sqrt{N})} \right) \, .
\end{equation}
The 2-Sample case uses the same formula as above, but with the slight variation where
\begin{equation} \label{eq:5}
N = \frac{n_1n_2}{n_1+n_2}\, .
\end{equation}
In both cases, $r$ is defined by
\begin{equation} \label{eq:6}
r = \frac{\sum_{i}^{}(x_i-\bar{x})(y_i-\bar{y})}{\sqrt{\sum_{i}^{}(x_i-\bar{x})^2}\sqrt{\sum_{i}^{}(y_i-\bar{y})^2}}\, .
\end{equation}

\RBnote{In general, treat the equations as part of the sentences.  This means punctuating them -- see the examples above.}
%use \verb|\,.| if the sentence ends with the equation, \verb|\,,| or \verb|\,;| if it's just the end of a clause.  The \verb|\,| puts a little space before the punctuation to make it more readable; if you need more space, use \verb|\;| or \verb|$\quad$|.}

% The basic Poisson regression model for count data is a special case of the GLM
% framework \cite{McCullagh+Nelder:1989}. It describes the dependence of a count
% response variable $y_i$ ($i = 1, \dots, n$) by assuming a Poisson distribution
% $y_i \sim \mathrm{Pois}(\mu_i)$. The dependence of the conditional mean
% $\E[y_i \, | \, x_i] = \mu_i$ on the regressors $x_i$ is then specified via a
% log link and a linear predictor
% %
% \begin{equation} \label{eq:mean}
% \log(\mu_i) \quad = \quad x_i^\top \beta,
% \end{equation}
% %
% where the regression coefficients $\beta$ are estimated by maximum likelihood
% (ML) using the iterative weighted least squares (IWLS) algorithm.
%
% \begin{leftbar}
% Note that around the \verb|{equation}| above there should be no spaces (avoided
% in the {\LaTeX} code by \verb|%| lines) so that ``normal'' spacing is used and
% not a new paragraph started.
% \end{leftbar}
%
% \proglang{R} provides a very flexible implementation of the general GLM
% framework in the function \fct{glm} \citep{Chambers+Hastie:1992} in the
% \pkg{stats} package. Its most important arguments are
% \begin{Code}
% glm(formula, data, subset, na.action, weights, offset,
%   family = gaussian, start = NULL, control = glm.control(...),
%   model = TRUE, y = TRUE, x = FALSE, ...)
% \end{Code}
% where \code{formula} plus \code{data} is the now standard way of specifying
% regression relationships in \proglang{R}/\proglang{S} introduced in
% \cite{Chambers+Hastie:1992}. The remaining arguments in the first line
% (\code{subset}, \code{na.action}, \code{weights}, and \code{offset}) are also
% standard  for setting up formula-based regression models in
% \proglang{R}/\proglang{S}. The arguments in the second line control aspects
% specific to GLMs while the arguments in the last line specify which components
% are returned in the fitted model object (of class \class{glm} which inherits
% from \class{lm}). For further arguments to \fct{glm} (including alternative
% specifications of starting values) see \code{?glm}. For estimating a Poisson
% model \code{family = poisson} has to be specified.
%
% \begin{leftbar}
% As the synopsis above is a code listing that is not meant to be executed,
% one can use either the dedicated \verb|{Code}| environment or a simple
% \verb|{verbatim}| environment for this. Again, spaces before and after should be
% avoided.
%
% Finally, there might be a reference to a \verb|{table}| such as
% Table~\ref{tab:overview}. Usually, these are placed at the top of the page
% (\verb|[t!]|), centered (\verb|\centering|), with a caption below the table,
% column headers and captions in sentence style, and if possible avoiding vertical
% lines.
% \end{leftbar}
%
% \begin{table}[t!]
% \centering
% \begin{tabular}{lllp{7.4cm}}
% \hline
% Type           & Distribution & Method   & Description \\ \hline
% GLM            & Poisson      & ML       & Poisson regression: classical GLM,
%                                            estimated by maximum likelihood (ML) \\
%                &              & Quasi    & ``Quasi-Poisson regression'':
%                                            same mean function, estimated by
%                                            quasi-ML (QML) or equivalently
%                                            generalized estimating equations (GEE),
%                                            inference adjustment via estimated
%                                            dispersion parameter \\
%                &              & Adjusted & ``Adjusted Poisson regression'':
%                                            same mean function, estimated by
%                                            QML/GEE, inference adjustment via
%                                            sandwich covariances\\
%                & NB           & ML       & NB regression: extended GLM,
%                                            estimated by ML including additional
%                                            shape parameter \\ \hline
% Zero-augmented & Poisson      & ML       & Zero-inflated Poisson (ZIP),
%                                            hurdle Poisson \\
%                & NB           & ML       & Zero-inflated NB (ZINB),
%                                            hurdle NB \\ \hline
% \end{tabular}
% \caption{\label{tab:overview} Overview of various count regression models. The
% table is usually placed at the top of the page (\texttt{[t!]}), centered
% (\texttt{centering}), has a caption below the table, column headers and captions
% are in sentence style, and if possible vertical lines should be avoided.}
% \end{table}

%% -- Illustrations ------------------------------------------------------------

%% - Virtually all JSS manuscripts list source code along with the generated
%%   output. The style files provide dedicated environments for this.
%% - In R, the environments {Sinput} and {Soutput} - as produced by Sweave() or
%%   or knitr using the render_sweave() hook - are used (without the need to
%%   load Sweave.sty).
%% - Equivalently, {CodeInput} and {CodeOutput} can be used.
%% - The code input should use "the usual" command prompt in the respective
%%   software system.
%% - For R code, the prompt "R> " should be used with "+  " as the
%%   continuation prompt.
%% - Comments within the code chunks should be avoided - these should be made
%%   within the regular LaTeX text.

@@

 Fasano--Franceschini Test (1987)
 Data:  sample1Data and sample2Data
 D-stat =  0.14 , p-value =  0.62
 Run Time (s) =  12.59964
\end{CodeOutput}
\end{CodeChunk}

Note, bootstrapping may significantly increase run time. To improve run time, one may adjust the \code{cores} parameter; see the R \pkg{parallel} package and the \fct{mclapply} function for further details.

>>>>

%% - \Abstract{} almost as usual
\Abstract{
\RBnote{Perhaps briefly state what the KS test is.}
The versatility of the univariate Kolmogorov-Smirnov (KS) test has
made it a cornerstone of statistical analysis and is commonly used
across the scientific disciplines. However, the test proposed by
Kolmogorov and Smirnov does not naturally extend to multidimensional
distributions.  Here, we present the \pkg{fasano.franceschini.test}
package, an  \proglang{R} implementation of the 2-D KS two--sample
test as defined by Fasano and Franceschini~\citep{Fasano1987}. The
\pkg{fasano.franceschini.test} package provides three improvements
over the current 2-D KS test on the Comprehensive \proglang{R} Archive
Network (CRAN): (i) the Fasano and Franceschini test has been shown to
run in $O(n^2)$ versus the Peacock implementation which runs in
$O(n^3)$; (ii) the package implements a procedure for handling ties in
the data; and (iii) the package implements a parallelized
bootstrapping procedure for improved significance testing. Ultimately,
the \pkg{fasano.franceschini.test} package presents a robust
statistical test for analysing random samples defined in 2-dimensions.
}

@@

% R -> Peacock test (computationally inificent)
% Python <- Implementation
% C <- Implementation
The Kolmogorov--Smirnov (KS) is a non--parametric, univariate
statistical test designed to assess whether a set of data is
consistent with a given probability distribution (or, in the two-sample 
case,  whether the two 
samples come from the same underlying distribution). First derived by
Kolmogorov and Smirnov in a series of
papers~\citep{Kolmogorov1933,Kolmogorov1933a,Smirnov1936,Smirnov1937,Smirnov1939,Smirnov1944,Smirnov1948},
the one-sample KS test defines the distribution of the quantity
$D_{KS}$, the maximal absolute difference between the empirical
cumulative distribution function (CDF) of a set of values and a
reference probability distribution. Kolmogorov and Smirnov's key
insight was proving the distribution of $D_{KS}$ was independent of
the CDFs being tested. Thus, the test can effectively be used to
compare any univariate empirical data distribution to any continuous
univariate reference distribution. The two-sample KS test could further
be used to compare any two univariate empirical data distributions
against each other to determine if they are drawn from the same
underlying univariate distribution.

The nonparametric versatility of the univariate KS test has made it a cornerstone of
statistical analysis and is commonly used across the scientific
disciplines~\citep{Atasoy2017,Chiang2018,Hahne2018,Hargreaves2020,Wong2020,Kaczanowska2021}.
However, the KS test as proposed by Kolmogorov
and Smirnov does not naturally extend to distributions in more
than one dimension. Fortunately, a solution to the dimensionality
issue was articulated by Peacock~\citep{Peacock1983} and later
extended by Fasano and Franceschini~\citep{Fasano1987}.

Currently, only the Peacock implementation of the 2-D two-sample KS
test is available in \proglang{R}~\citep{R} with the
\pkg{Peacock.test} package via the \fct{peacock2} function, but this
has been shown to be markedly slower than the Fasano and Franceschini
algorithm~\citep{Lopes2007}. A \proglang{C} implementation of the
Fasano--Franceschini test is available in~\cite{numericalRecipes};
however, arguments have been made to the validity of the
implementation of the test not being
distribution-free~\citep{Babu2006}. Furthermore, the \proglang{C}
implementation statistical testing is based on a fit to Monte Carlo
Simulation and is only valid for significance levels $\alpha \lessapprox 0.20$, 
which may present potential issues in some analysis.

Here we present the \pkg{fasano.franceschini.test} package as an
\proglang{R} implementation of the 2-D two-sample KS test described
by Fasano and Franceschini~\citep{Fasano1987}. The
\pkg{fasano.franceschini.test} package provides two improvements over
the current 2-D KS test on the Comprehensive \proglang{R} Archive
Network (CRAN): (i) the Fasano and Franceschini test has been shown to
run in $O(n^2)$ versus the Peacock implementation which runs in
$O(n^3)$; and (ii) the package implements a bootstrapping procedure
for improved significance testing and mitigates the limitations of
the test brought noted by~\citep{Babu2006}.

@@

The Kolmogorov--Smirnov (KS) test is a non--parametric method for
determining whether a sample is consistent with a given probability
distribution~\citep{Stephens1992a}.
%While a comprehensive discussion pertaining the intricacies of the 1-D KS test can be found elsewhere~\citep{Stephens1992a}, we will present a brief overview as motivation for the basis and intuition of the 2-D KS test.
%
%As described previously, the 1-D KS test can be divided into two variants of the same statistical procedure:
%
%\begin{leftbar}
%\textbf{The One-Sample Test}
%\begin{itemize}
%\item Compares the data to analytical distribution.
%\end{itemize}
%\textbf{The Two-Sample Test}
%\begin{itemize}
%\item Compares two data samples to determine if they are drawn from the same underlying probability distribution.
%\end{itemize}
%\end{leftbar}
%
%In both cases, the KS test works as follows:
%
In one dimension, the Kolmogorov-Smirnov statistic ($D_{KS}$) is the
defined by the maximum absolute difference between the cumulative
density functions of the data and model (one--sample), or between the
two data sets (two--sample), as illustrated in
\textbf{Figure~\ref{fig:kstest1D}}.
%. In the following example, the orange and blue curve in the left panel represents the underlying distribution (pdf) for sample1 and sample2, respectively (\textbf{Figure \ref{fig:kstest1D}}). The eCDF is then defined by data drawn from these underlying distribution, and the D--stat is the computed as the largest between difference between the eCDF (black dotted line) in the right panel.

\begin{figure}[hbt]
\centering
\includegraphics{pdfvsCDF}
\caption{\label{fig:kstest1D} \textbf{LEFT:} Probability density function (PDF) of two normal distributions: orange sample~1, $\mathcal{N}(\mu = 0,\,\sigma^{2} = 1)$; blue sample~2, $\mathcal{N}(\mu = 5,\,\sigma^{2} = 1)$. \textbf{RIGHT:} Cumulative density functions (CDF) of the two PDFs; the black dotted line represents the maximal absolute difference between the CDFs ($D_{KS}$).
}
\end{figure}

@@

In the large--sample limit ($n \geq 80$), it can be shown~\citep{Kendall1946} that $D_{KS}$ converges in distribution to
\begin{equation} \label{eq:1}
\Phi(\lambda) = 2 \sum_{k=1}^{\infty} -1^{k-1}e^{-2k^2\lambda^2} \,.
\end{equation}
%Given the maximum absolute difference ($D_{KS}$) and \textbf{Equation~\ref{eq:1}}, one can  compute a $p$-value for  hypothesis testing to determine if there is a significance difference between the cumulative density functions of the data and model (one-sample), or between the two data sets (two-sample).
%
%\textbf{For the 1-Sample case}, the $p$-value probability of a given D is defined by
In the one-sample case with a sample of size $n$, the $p$~value is given by
\begin{equation} \label{eq:2}
\mathbb{P}(D > observed) = \Phi ( D\sqrt{n})\,;
\end{equation}
in the two-sample case, the $p$~value is given by
%where $N$ is the number of data points sampled. A $p$-value $< 0.05$, indicates a significant difference between the samples.
%
%\textbf{In the 2-Sample case}, the $p$-value probability of a given D is defined by
\begin{equation} \label{eq:3}
\mathbb{P}(D > observed) = \Phi \left( D\sqrt{\frac{n_1n_2}{n_1+n_2}} \right)\,.
\end{equation}
%
where $n_1$ and $n_2$ are the number of observations in the first and second samples respectively. 
%A $p$-value $< 0.05$, indicates a significant difference between the samples.

\subsection{Higher dimensional variations: Peacock Test (1983) and Fasano--Franceschini Test (1987)}
%Similar to the 1-D KS test, the obvious solution to extend the theory into higher dimensional would be to define the maximal cumulative difference between the two 2-D distributions. This however, presents an issue as the cumulative distribution function is not well defined in more than one dimension. In 2-D, there are 4 ways (3 independent ways) of defining the cumulative distribution, since the direction in which we define choose to order the points in the x and y are arbitrary (\textbf{Figure \ref{fig:kstest2Dissue}}). In N dimensional space there are $2^{N}-1$ independent ways of defining the cumulative distribution function~\citep{Peacock1983}.
Extending the above to two or higher dimension is complicated by the
fact that CDFs are not well-defined in more than one dimension.  In
2-D, there are 4 ways (3 independent ways) of defining the cumulative
distribution, since the direction in which we order the $x$ and $y$
points are arbitrary (\textbf{Figure \ref{fig:kstest2Dissue}}); more
generally, in $k$-dimensional space there are $2^{k}-1$ independent
ways of defining the cumulative distribution
function~\citep{Peacock1983}.

\begin{figure}[t!]
\centering
\includegraphics{CDF2Dissue}
\caption{\label{fig:kstest2Dissue} Four ways (3 independent) of defining the cumulative distribution for a given point in 2-D.  Here, the orange point $(X,Y)$ is chosen as the origin; the density of observations may be integrated as 
$\mathbb{P}(x < X, y > Y)$ (A); 
$\mathbb{P}(x < X \cup y < Y)$ (B); 
$\mathbb{P}(x < X, y < Y)$ (C); 
$\mathbb{P}(x > X, y > Y)$ (D). 
%$P(x < X, y < Y) + P(x < X, y > Y) + P(x > X, y < Y) = 1 - P(x > X, y > Y)$
}
\end{figure}

\cite{Peacock1983} solved the higher dimensionality issue by defining
the 2-D test statistic as the largest difference between the empirical
and theoretical cumulative distributions, after taking all possible
ordering combinations into account. Peacock's test thus computes the
total integrated probability---i.e.\,fraction of data---tin each of
the four quadrants around all possible points in the data. For
example, for a given $n$ points in a two-dimensional space, the
cumulative distribution functions is calculated in the $4n^2$
quadrants of the plane defined by all pairs ($X_i$ ,$Y_j$), $X_i$ and
$Y_j$ being coordinates of any pairs of points in the given sample. By
ranging over all possible pairs of data points and quadrants, the
2-dimensional $D$ statistic is defined by the maximal difference of
the integrated probabilities between samples.

The variation defined by \cite{Fasano1987}  was to only consider
quadrants centered in each point of the given samples to compute the
cumulative distribution functions.  Rather than looking over all 
$X_i, Y_j: i,j={1,n}$, Fasano and Franceschini only used the detected points
$X_i, Y_i : i={1,n}$. 
\RBnote{I don't follow the preceding line? (Notation?)}
Thus for any given $n$ points in a two-dimensional space, those $n$ points define $4n$ (rather than
$4n^2$) quadrants.  The proceedure is illustrated in \textbf{Figure~\ref{fig:kstest2D}}.
The algorithm loops through each point in one sample in turn to define
the origin of 4 quadrants (grey dotted lines in
\textbf{Figure~\ref{fig:kstest2D}}). The fraction of points in each
sample is computed in each quadrant, and the quadrant with the maximal
difference is designated with the current maximum for the specified
origin. By iterating over all data points and quadrants, the test
statistic $D_{FF,1}$ is defined by the maximal difference of the
integrated probabilities between samples in any quadrant for any
origin from the first sample.  In \textbf{Figure~\ref{fig:kstest2D}},
using the orange point as the origin, the maximal difference is
$D_{FF,1} = 0.52$.

@@

%\RBnote{Since $D_{FF}$ from one sample isn't the final $D_{FF}$, you should note it appropriately (eg, $D_{FF,1}$ and $D_{FF,2}$). - ENC: Text has been updated accordingly}
This process is repeated using the points from \textit{other} sample
as the origins to compute the maximal $D_{FF,2}$ with origins from the
second sample. $D_{FF,1}$ and $D_{FF,2}$  are then averaged to compute
the overall $D_{FF}$ for hypothesis testing. \RBnote{clarify: is
$D_{FF}=(D_{FF,1}+D_{FF,2})/2$, or is it weighted by the $n$?}

It may be that some points are tied with the $X$ and/or $Y$
coordinates of the origin, creating an ambiguity when computing the
fraction of points in each quadrant. Since the 
test attempts to define the maximal difference of the cumulative
probabilities, a natural solution would be to treat a point that is
tied with the current $X$ and/or $Y$ coordinates of the origin as equally
likely to have been drawn from any of the tied quadrants. 
Hence, any data point
sharing the same $X$ or $Y$ coordinate as the origin is evenly
distributed across the tied quadrants, with each of the two quadrants
recieving half a count. Any data point sharing the both the same $X$ and $Y$
coordinates as the current origin (including the origin itself) is evenly
distributed across all quadrants, with all four quadrants recieveing a
quarter count.

@@

Using Monte Carlo simulation, Fasano and Franceschini created a
look-up table of critical values of $D_{FF}$ as a function of
$D_{FF}$, the sample size, and the coefficient of correlation $r$.
\cite{numericalRecipes} later defined an approximate fit to the lookup table
as follows. 
%
For the a single sample of size $n$,
\begin{equation} \label{eq:4}
\mathbb{P}(d_{FF} > D_{FF}) = \Phi \left( \frac{D_{FF}\sqrt{n}}{1+\sqrt{1-r^2}(0.25-0.75/\sqrt{n})} \right) \, .
\end{equation}
\RBnote{Is $\Phi$ the same function defined in Eq~\ref{eq:1}?}
The two sample case uses the same formula as above, but with the slight variation where
\begin{equation} \label{eq:5}
n = \frac{n_1n_2}{n_1+n_2}\, .
\end{equation}
In both cases, $r$ is defined in the usual way as
\begin{equation} \label{eq:6}
r = \frac{\sum_{i}(X_i-\bar{X})(Y_i-\bar{Y})}{\sqrt{\sum_{i}(X_i-\bar{X})^2}\sqrt{\sum_{i}(Y_i-\bar{Y})^2}}\, .
\end{equation}

@@

 Fasano--Franceschini Test (1987)
 Data:  sample1Data and sample2Data
 D-stat =  0.14 , p-value =  0.62
 Run Time (s) =  12.59964
\end{CodeOutput}
\end{CodeChunk}
To improve run time, one may adjust the \code{cores} parameter; see the R \pkg{parallel} package and the \fct{mclapply} function for further details.
